Insert your title here

### 2023-07-20 manuscript_JQAS/manuscript_JQAS.Rnw
Some changes

All of the previous work mentioned here focuses on change point analysis in {\bf univariate} time series. However, recent methodolgical developments in change point analysis allow for change point analysis in {\bf multivariate} time series, which is the focus of this current manuscript.  This work leverages techniques such as the Double CUSUM Binary Segmentation algorithm (\cite{Cho2016}) and the Sparsified Binary Segmentation algorithm (\cite{ChoFryzlewwicz2014}) to look for change points in Major League Baseball at two main levels.  First, we look for structural changes in the league as a whole across teams to empirically define different eras in baseball.  Second, we look for change points within a team to determine eras of a team.  This second analysis can be used to identify the beginning and end of so called "dynasties", periods of sustained excellent performance by a team.  In addition, we can also identify the opposite, sustained periods of poor performance.

@@

% <!-- Just Notes:  -->
% <!-- Palacios-Huerta I. (2004). Structural changes during a Century of the World’s most popular sport. Statistical Methods and Applications, 13, 241–258 -->
% <!-- @PalaciosHuerta2004 Structural Breaks in soccer -->
% 
% 
% <!-- @Nieswiadomy2012 Was there a structural break in Barry Bonds’s Bat?  Looking for structural breaks in Bonds OPS times series. Use the uni root test on monthly data from 86-07.  The identify break points in June 1993 and Sept 2000.  -->
% 
% 
% <!-- In contrast to their methods, we use  more modern methods Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewwicz2014). -->
% 
% <!-- https://onlinelibrary.wiley.com/doi/epdf/10.1111/j.1465-7295.2007.00026.x -->
% <!-- Fort and Lee 2007: Looks for structural breaks in NBA, NHL, and NFL -->
% 
% 
% 
% <!-- https://www.baseball-reference.com/bullpen/History_of_baseball_in_the_United_States -->
% 
% <!-- https://thesportjournal.org/article/examining-perceptions-of-baseballs-eras/ -->

@@

% <!-- univariate time series of the means and SD's of several statistics (i.e. SLUG, HR rate, RBI rate, and batting average) to look for structural breaks via the Lagrange Multiplier (LM) unit root test @LeeandStrazicich2003.  Notably they find structural breaks in slugging percentage in 1921 and 1992.  In contrast to their methods, we use  more modern methods Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewwicz2014). -->

@@

% <!-- @Groothius2017 is the most closely related work to ours.  They use univariate time series of the means and SD's of several statistics (i.e. SLUG, HR rate, RBI rate, and batting average) to look for structural breaks via the Lagrane Multiplier (LM) unit root test @LeeandStrazicich2003.  Notably they find structural breaks in slugging percentage in 1921 and 1992.  In contrast to their methods, we use  more modern methods Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewwicz2014).   -->

@@

\section{Methods}
%@ChoFryzlewwicz2014 and @Cho2016

@@

To determine the testing criterion $\pi_{n,T}^\phi$, a Generalized Dynamic Factor Model (GDFM) bootstrapping algorithm is employed (CITE XX), accounting for potential correlations within and between high-dimensional time series. This methodology enables the detection of second-order change points, utilizing methodology from \cite{ChoFryzlewwicz2014} Haar wavelet periodograms and cross-periodograms, as opposed to the traditional CUSUM statistic. For further details, refer to \cite{Cho2016} and \cite{ChoFryzlewwicz2014}.  These methods were implemented using the \texttt{hdbinseg} \cite{hdbinseg} in the R programming lanugage (\cite{R2023language}).

@@

\section{Data}
We obtained multi-year baseball statistics from the Lahman database (\cite{Friendly2021Lahman}). Over the course of the sport's history, numerous statistics have been collected, and for our analysis, we focused on year-end statistics pertaining to the teams. To ensure consistency in our methodology, it was necessary for each team to have existed from 1900 until the end of the dataset in 2020 (IS THIS BECAUSE THE CHO2016 method requires the time series to be the same length?). This timeframe of 120 years, denoted as T, encompassed a total of 16 franchise teams, namely NYY, BOS, LAD, ATL, CHW, CHC, CIN, CLE, DET, BAL, SFG, OAK, PHI, PIT, STL, and MIN. The statistics of interest for these teams include runs, hits, home runs, balls, strikeouts, at-bats, stolen bases, number of games played in a season, attendance, runs against, hits against, home runs against, balls against, and strikeouts against. However, a few statistics, such as balls, stolen bases, strikeouts, and attendance, had missing data for certain years. To address this, we employed multiple imputations using classification and regression trees (\cite{Vanbuuren1999flexible}) to fill in the missing values and ensure consistent data across all analyses.  (HOW DID WE DO THRE IMPUTATION?  Was is single imputation?  )

\section{Results}
\subsection{Eras}
Since we are interested in finding where in time these different eras of baseball have occurred, we first examine each baseball statistic of interest as their own analysis for all 16 teams. Those baseball statistics are home runs, strikeouts, balls, stolen bases, runs, and attendance which were all standardized to the number of games played in a season since we are interested in seeing how the different statistics change from season to season. Using the methodology from Cho and Fry (CITE) a thresholding parameter was determined from bootstrapping and then used to find where change points in mean and variance existed in the panel of teams data of a given baseball statistic. To examine the idea of changes throughout teams and types of statistics (initially), for each season the average baseball statistic was found from all the teams that year and then an average time series was created of each of the types of statistics of interest. Then the change point methodology was performed on those 6 time series of average statistic types.

@@

\section{Conclusions}

>>>>

All of the previous work mentioned here focuses on change point analysis in {\bf univariate} time series. However, recent methodolgical developments in change point analysis allow for change point analysis in {\bf multivariate} time series, which is the focus of this current manuscript.  This work uses the Double CUSUM Binary Segmentation algorithm (\cite{Cho2016,ChoFryzlewicz2014}) to look for change points in Major League Baseball at two main levels.  First, we look for structural changes in the league as a whole across teams to empirically define different eras in baseball.  Second, we look for change points within a team to determine eras of a team.  This second analysis can be used to identify the beginning and end of so called "dynasties", periods of sustained excellent performance by a team.  In addition, we can also identify the opposite, sustained periods of poor performance.

@@

% <!-- Just Notes:  -->
% <!-- Palacios-Huerta I. (2004). Structural changes during a Century of the World’s most popular sport. Statistical Methods and Applications, 13, 241–258 -->
% <!-- @PalaciosHuerta2004 Structural Breaks in soccer -->
% 
% 
% <!-- @Nieswiadomy2012 Was there a structural break in Barry Bonds’s Bat?  Looking for structural breaks in Bonds OPS times series. Use the uni root test on monthly data from 86-07.  The identify break points in June 1993 and Sept 2000.  -->
% 
% 
% <!-- In contrast to their methods, we use  more modern methods Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewicz2014). -->
% 
% <!-- https://onlinelibrary.wiley.com/doi/epdf/10.1111/j.1465-7295.2007.00026.x -->
% <!-- Fort and Lee 2007: Looks for structural breaks in NBA, NHL, and NFL -->
% 
% 
% 
% <!-- https://www.baseball-reference.com/bullpen/History_of_baseball_in_the_United_States -->
% 
% <!-- https://thesportjournal.org/article/examining-perceptions-of-baseballs-eras/ -->

@@

% <!-- univariate time series of the means and SD's of several statistics (i.e. SLUG, HR rate, RBI rate, and batting average) to look for structural breaks via the Lagrange Multiplier (LM) unit root test @LeeandStrazicich2003.  Notably they find structural breaks in slugging percentage in 1921 and 1992.  In contrast to their methods, we use  more modern methods Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewicz2014). -->

@@

% <!-- @Groothius2017 is the most closely related work to ours.  They use univariate time series of the means and SD's of several statistics (i.e. SLUG, HR rate, RBI rate, and batting average) to look for structural breaks via the Lagrane Multiplier (LM) unit root test @LeeandStrazicich2003.  Notably they find structural breaks in slugging percentage in 1921 and 1992.  In contrast to their methods, we use  more modern methods Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewicz2014).   -->

@@

\section{Data}
%EXPLAIN DATA NEEDS FOR 1) ERAS and 2) DYNASTIES.
We obtained multi-year baseball statistics from the Lahman database (\cite{Friendly2021Lahman}). Over the course of the sport's history, numerous statistics have been collected, and for our analysis, we focused on year-end statistics pertaining to the teams. In order to implement the DC algorithm, each time series is required to be the same length.  As a result, when studying eras in our analysis we focus on a set of 16 teams that have all existed over the period from 1900 to 2020 because this period of time covers most of the history of organized professional baseball and there are enough teams that have all existed over this time period to perform meaningful analysis.  If we had instead chosen to extend this period to, for instance, 1890 to 2020, we would only have had XX teams available.   This timeframe of $T=120$ years encompassed a total of 16 franchise teams listed in table \ref{tab:120year:franch}. The statistics of interest for these teams include runs, hits, home runs, balls, strikeouts, at-bats, stolen bases, number of games played in a season, attendance, runs against, hits against, home runs against, balls against, and strikeouts against. However, a few statistics, such as balls, stolen bases, strikeouts, and attendance, had missing data for certain years. To address this, we employed multiple imputations using classification and regression trees (\cite{Vanbuuren1999flexible}) to fill in the missing values and ensure consistent data across all analyses.  (HOW DID WE DO THRE IMPUTATION?  Was is single imputation?  )

For the dynasty analysis we used teams over different periods of time for every year that each team has existed....MORE DETAILS.

\begin{table}
\begin{tabular}{|r|}
Franchise  \\
\hline
NYY Full names 
 NYY, BOS, LAD, ATL, CHW, CHC, CIN, CLE, DET, BAL, SFG, OAK, PHI, PIT, STL, and MIN\\
\hline
\end{tabular}
\caption{}
\label{tab:120year:franch}
\end{table}

\section{Methods}
%@ChoFryzlewicz2014 and @Cho2016

@@

To determine the testing criterion $\pi_{n,T}^\phi$, a Generalized Dynamic Factor Model (GDFM) bootstrapping algorithm is employed (CITE XX), accounting for potential correlations within and between high-dimensional time series. This methodology enables the detection of second-order change points, utilizing methodology from \cite{ChoFryzlewicz2014} Haar wavelet periodograms and cross-periodograms, as opposed to the traditional CUSUM statistic. For further details, refer to \cite{Cho2016} and \cite{ChoFryzlewicz2014}.  These methods were implemented using the \texttt{hdbinseg} \cite{hdbinseg} in the R programming language (\cite{R2023language}).

Discuss somewhere how we can find mean change points and ALSO variance change points.

SEGUE parageraph Using eras and dynasties:::::::::

Since we are interested in finding where in time these different eras of baseball have occurred, we first examine each baseball statistic of interest as their own analysis for all 16 teams. Those baseball statistics are home runs, strikeouts, balls, stolen bases, runs, and attendance which were all standardized to the number of games played in a season since we are interested in seeing how the different statistics change from season to season. Using the methodology from \cite{ChoFryzlewicz2014} a thresholding parameter was determined from bootstrapping and then used to find where change points in mean and variance existed in the panel of teams data of a given baseball statistic. To examine the idea of changes throughout teams and types of statistics (initially), for each season the average baseball statistic was found from all the teams that year and then an average time series was created of each of the types of statistics of interest. Then the change point methodology was performed on those 6 time series of average statistic types.

@@

\section{Results}
\subsection{Eras}
Blah blah 
\subsection{Dynasties}
Blah blah

@@

\section{Conclusions}
Baseball researchers
